Decompositions of congruence subgroups \\ of Chevalley groups

### 2018-01-30 paper.tex
Greatly improved readability of the proof of Theorem 1

Let $J\subseteq \Pi$ be a subset of fundamental roots. Define
\begin{align*}
& \shape(J, \beta)=\sum\nolimits_{i\in J} m_i(\beta) \alpha_i, \\
& \Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\}, \\
& \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha) \in \mathbb{Z}_{\geqslant0} \Phi^\pm \}, \\
& S_J^\pm = \Delta_J \sqcup \Sigma_J^\pm.
\end{align*}
Clearly, $\Delta_J$ is a reductive subset, while $S^\pm_J$ and $\Sigma^\pm_J$ are parabolic and special subsets, respectively.
For two disjoint subsets $I, J\subseteq \Pi$ one has 
\[ \Sigma^\pm_{I \cup J} = \Sigma^\pm_I\cup\Sigma^\pm_J, \quad \Delta_{I\cup J} = \Delta_I \cap \Delta_J. \]
We omit curly braces in the above notations when $J$ is a one- or two-element set, e.\,g. $\Delta_k=\Delta_{\{k\}}$ and $\Sigma_{i, j}^\pm=\Sigma_{\{i, j\}}^\pm$, etc. We call $\shape(\{s\}, \beta)$ the $s$-shape of $\beta$.

@@

Denote by $N_{rs}$ the submonoid of $\E(\Phi, R)$, consisting of the elements which normalize the subset $A_{rs}$, i.\,e.
\[ N_{rs} = \left\{ g\in \E(\Phi, R)\ \middle|\ g \cdot A_{rs} \cdot g^{-1} \subseteq A_{rs} \right\}. \]

\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup (\Phi^+ \setminus (\Sigma^+_r \cap \Sigma^+_s))$ one has $X_\alpha(R) \subseteq N_{rs}$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\U(\Sigma_r^- \cap \Sigma_s^-, I)$ is normalized by $\E(\Delta_{\{r, s\}}, R)$, we obtain the result for $\alpha \in \Delta_{r, s}$.

Now if $\alpha$ lies in $\Phi^+ \setminus (\Sigma_r^+ \cap \Sigma_s^+)$, it is contained either in $\Delta_r$ or $\Delta_s$.
Assume, for example, the latter. By the Levi decomposition we have $\U(\Sigma_r^- \cap \Sigma_s^-, I)^{X_\alpha(R)} \subseteq \U(\Sigma_s^-, I) \subseteq A_{rs}$,
which again implies the assertion of the lemma.
\end{proof}

@@

Denote by $L_{rs}$ the set consisting of elements $g\in \E(\Phi, R, I)$ such that $g \cdot A_{rs} \subseteq A_{rs}$.
It is easy to see that $L_{rs}$ contains $\EP_r(R, I)$ and is normalized by $N_{rs}$, i.\,e. ${L_{rs}}^{N\ssub{rs}}\subseteq L_{rs}$. Indeed, for $g\in N_{rs}$, $h\in L_{rs}$ one has
\begin{equation}\label{rel:NnormL} h^g \cdot A_{rs} = g^{-1} \cdot h \cdot g \cdot A_{rs} \subseteq g^{-1} \cdot h \cdot A_{rs} \cdot g \subseteq {A_{rs}}^g \subseteq A_{rs}.\end{equation}

The first step of the proof is to show that $N_{rs}$ contains sufficiently many elements. 
This is accomplished in~\cref{lemma:Stein_reduction} below, where we invoke the stability conditions for the first time.
We show that $N_{rs}$ contains the root subgroup $X_{-\alpha\ssub{r}}(R)$, which together with \cref{lemma:dv-normal} and \cref{item-egen} immediately implies that $N_{rs}$ contains the Levi subgroup $\E(\Delta_s, R)$ and hence the extended Weyl group $\widetilde{W}(\Delta_s)$.

Our next goal is to demonstrate the inclusion $\U(\Sigma_s^+, R) \subseteq N_{rs}$ (which would imply $\EP_s(R) \subseteq N_{rs}$).
All the possibilities for $\Phi$, $s$ and $r$ fall into one of the following three cases:
\begin{enumerate}
 \item \textit{Case $m_s(\alpha_{\mathrm{max}})=1$ and $\Phi$ is simply laced.}
  In this case all the roots of $\Sigma^+_s$ have the same $s$-shape and length, hence the assertion follows from \cref{item-trans2}, \eqref{rel:NnormL} and the fact that we already have the inclusion $X_{\alpha_s}(R) \subseteq N_{rs}$.
 \item \textit{Case $\Phi = \rB_\ell$.} Using the argument from the previous case, we get that $N_{rs}$ contains the subgroup generated by the root subgroups $X_\alpha(R)$, where $\alpha$ varies over the long roots of $\Sigma_s^+$. 
  It remains to prove the inclusion $X_{1, 0}(R) \subset N_{rs}$.
  From Chevalley commutator formula we obtain
\begin{equation*} \label{rel:CCF-specBC}
x_{1, 0}(ab) = [x_{1, 2}(a), x_{2, 0}(b)] \cdot x_{1, -2}(-a b^2).
\end{equation*}
  Since $X_{2, 0}(R)$, $X_{1, -2}(R)$ and $X_{1,2}(R)$ are contained in $N_{rs}$, the inclusion $X_{1, 0}(R) \subseteq N_{rs}$ follows.
 \item \textit{Case $\Phi = \rC_\ell$.}
 We settle the remaining case in \cref{lemma:DVST}, where we invoke the stability conditions one more time. 
 In fact, we prove a stronger assertion, namely that $N_{rs}$ contains $X_{-\alpha\ssub{s}}(R)$ and, consequently, that $N_{rs} = \E(\Phi, R)$.
\end{enumerate}

In view of \cite[Theorem~3.4]{S} to prove the theorem it suffices to show that $L_{rs}$ contains the generating set $\mathcal{Z}(\Sigma_s^-)$ of $\E(\Phi, R, I)$.
From the inclusion $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)}$ and \eqref{rel:NnormL} it follows that we only need to verify that $\mathcal{X} \subseteq L_{rs}$.
In turn, the last inclusion follows from the same transitivity argument as above if we use the fact that $X_\alpha(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$ for $\alpha \in \Delta_r \cup \Sigma_r$.
For $\Phi=\rB_\ell$ we need to apply Chevalley commutator formula once again:
\[ X_{0, 1}(I) \subseteq X_{2, 1}(I) \cdot X_{2, 1}(I)^{x_{0, 2}(1)} \cdot X_{-2, 1}(I) \subseteq L_{rs} \cdot L_{rs}^{N_{rs}} \cdot L_{rs} \subseteq L_{rs}. \qedhere \] 
\end{proof}

>>>>

Let $J\subseteq \Pi$ be a subset of fundamental roots. Define
\begin{align*}
& \shape(J, \beta)=\sum\nolimits_{i\in J} m_i(\beta) \alpha_i, \\
& \Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\}, \\
& \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha) \in \mathbb{Z}_{\geqslant0} \Phi^\pm \}, \\
& S_J^\pm = \Delta_J \sqcup \Sigma_J^\pm.
\end{align*}
Clearly, $\Delta_J$ is a reductive subset, while $S^\pm_J$ and $\Sigma^\pm_J$ are parabolic and special subsets, respectively.
For two disjoint subsets $I, J\subseteq \Pi$ one has 
\[ \Sigma^\pm_{I \cup J} = \Sigma^\pm_I\cup\Sigma^\pm_J, \quad \Delta_{I\cup J} = \Delta_I \cap \Delta_J. \]
We omit curly braces in the above notations when $J$ is a one- or two-element set, e.\,g. $\Delta_k=\Delta_{\{k\}}$ and $\Sigma_{i, j}^\pm=\Sigma_{\{i, j\}}^\pm$, etc.
We call $\shape(\{s\}, \beta)$ simply the $s$-shape of $\beta$.

@@

Denote by $N_{rs}$ the subgroup of $\E(\Phi, R)$, consisting of elements normalizing the subset $A_{rs}$, i.\,e.
\[ N_{rs} = \left\{ g\in \E(\Phi, R)\ \middle|\ g \cdot A_{rs} \cdot g^{-1} \cup g^{-1} \cdot A_{rs} \cdot g \subseteq A_{rs}\right\}. \]

\begin{lemma}\label{lemma:dv-normal} 
For every $\alpha \in \Delta_{r, s} \cup \Phi^+ $ one has $X_\alpha(R) \subseteq N_{rs}$. \end{lemma}
\begin{proof}
Notice that for every $i$ the group $\EP_i(R, I)$ is normalized by $\EP_i(R)$, hence it is normalized by $X_\alpha(R)$, $\alpha \in S_i^+$.
Since $\E(\Delta_{r, s}, R)$ normalizes $\U(\Sigma_r^- \cap \Sigma_s^-, I)$, we obtain the assertion for $\alpha \in \Delta_{r, s}$.

Now let $\alpha$ be a positive simple root. It is clear that $\alpha$ lies either in $\Delta_r$ or $\Delta_s$.
Assume, for example, the latter. From the Levi decomposition we get $\U(\Sigma_r^- \cap \Sigma_s^-, I)^{X_\alpha(R)} \subseteq \U(\Sigma_s^-, I) \subseteq A_{rs}$,
consequently $X_{\alpha}(R) \subseteq N_{rs}$.
Now it is clear that $N_{rs}$ contains the subgroup $\U(\Phi^+, R)$ generated by $X_{\alpha}(R)$, $\alpha\in\Pi$, which implies the assertion.
%TODO: Why the positive unipotent radical is generated by simple root subgroups???
\end{proof}

@@

Denote by $L_{rs}$ the monoid consisting of elements $g\in \E(\Phi, R, I)$ such that $g \cdot A_{rs} \subseteq A_{rs}$.
It is easy to see that $L_{rs}$ contains $\EP_r(R, I)$ and is normalized by $N_{rs}$, i.\,e. ${L_{rs}}^{N\ssub{rs}}\subseteq L_{rs}$. Indeed, for $g\in N_{rs}$, $h\in L_{rs}$ one has
\begin{equation}\label{rel:NnormL} h^g \cdot A_{rs} = g^{-1} \cdot h \cdot g \cdot A_{rs} \cdot g^{-1} \cdot g \subseteq g^{-1} \cdot h \cdot A_{rs} \cdot g \subseteq {A_{rs}}^g \subseteq A_{rs}.\end{equation}

First of all notice that $N_{rs}$ is sufficiently large.
\cref{lemma:Stein_reduction} below guarantees that $N_{rs}$ contains the root subgroup $X_{-\alpha_{r}}(R)$, which together with~\cref{lemma:dv-normal} 
 implies that $\EP_s(R) \subseteq N_{rs}$ and consequently that $\EP_s(R, I) \subseteq L_{rs}$. Indeed for $g\in \EP_s(R, I) \subseteq \EP_s(R)$ one has
$g \cdot L_{rs} = g \cdot L_{rs} \cdot g^{-1} \cdot g \subseteq L_{rs} \cdot g \subseteq L_{rs}.$

%This is accomplished in~\cref{lemma:Stein_reduction} below, where we invoke the stability conditions for the first time.
%which together with \cref{lemma:dv-normal} implies that $N_{rs}$ contains the Levi subgroup $\E(\Delta_s, R)$.
% 
 %$\EP_s(R) \subseteq N_{rs}$

The next step of the proof is to show that $\U(\Sigma_s^-, I) \subseteq L_{rs}$.
All the possibilities for $\Phi$, $s$ and $r$ fall into one of the following three cases:
\begin{enumerate}
 \item \textit{Case $m_s(\alpha_{\mathrm{max}})=1$ and $\Phi$ is simply laced.}
  In this case all the roots of $\Sigma^-_s$ have the same $s$-shape and length.
  Since $\widetilde{W}(\Delta_s) \subseteq \E(\Delta_s, R) \subseteq N_{rs}$ and $X_{-\alpha_s}(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$,
   the required inclusion $\U(\Sigma_s^-, I) \subseteq L_{rs}$ follows from \cref{item-trans2}.  
 \item \textit{Case $\Phi = \rB_\ell$, $s=1$, $r=\ell$.} 
  The argument from the previous case shows that $L_{\ell,1}$ contains root subgroups $X_\alpha(I)$, where $\alpha$ varies over the long roots of $\Sigma_1^-$. 
  The inclusion $X_{0, 1}(I) \subseteq L_{\ell, 1}$ for the remaining short root follows from inclusions $X_{-1, 2}(I) \cup X_{2, 1}(I) \subseteq L_{\ell,1}$, $X_{2, 0}(R) \subseteq N_{\ell,1}$
  using Chevalley commutator formula $x_{0, 1}(-ab) = [x_{-1,2}(a),\, x_{2,0}(b)] \cdot x_{2,1}(ab^2).$
 \item \textit{Case $\Phi = \rC_\ell$, $s=1$, $r=\ell$.} 
  Similarly, one can show that $L_{\ell,1}$ contains $X_\alpha(I)$ for all short $\alpha \in \Sigma_1^-$.
  However, the inclusion $X_{-1, 1}(I) \subseteq L_{\ell, 1}$ for the remaining long root does not immediately follow from commutator formulae.
  To verify it we invoke stability conditions once again.
  From~\cref{lemma:DVST} we obtain that $N_{\ell, 1}$ contains $X_{-\alpha\ssub{1}}(R)$ and therefore coincides with the whole $\E(\Phi, R)$, which gives the required assertion.
\end{enumerate}

Thus, we have just shown that $\mathcal{X}\subseteq L_{rs}$. 
Recall from~\cite[Theorem~3.4]{S} that the group $\E(\Phi, R, I)$ is generated by $\mathcal{Z}(\Sigma_s^-)$.
On the other hand, from $\mathcal{Z}(\Sigma_s^-) \subseteq \mathcal{X}^{\EP_s(R)} \subseteq L_{rs}^{N_{rs}} \subseteq L_{rs}$ we conclude that 
 $\E(\Phi, R, I) \subseteq L_{rs}$, which implies the assertion of the theorem.
%Thus, to prove our assertion it suffices to show that $L_{rs}$ contains $\mathcal{Z}(\Sigma_s^-)$.
%From the inclusion  and \eqref{rel:NnormL} it follows that we only need to verify the inclusion $\mathcal{X} \subseteq L_{rs}$.
%In turn, the last inclusion follows from the same transitivity argument as above if we use the fact that $X_\alpha(I) \subseteq \EP_r(R, I) \subseteq L_{rs}$ for $\alpha \in \Delta_r \cup \Sigma_r$.
%For $\Phi=\rB_\ell$ we need to apply Chevalley commutator formula once again:
%\[ X_{0, 1}(I) \subseteq X_{2, 1}(I) \cdot X_{2, 1}(I)^{x_{0, 2}(1)} \cdot X_{-2, 1}(I) \subseteq L_{rs} \cdot L_{rs}^{N_{rs}} \cdot L_{rs} \subseteq L_{rs}. \qedhere \] 
\end{proof}
